---
title: |
  Distance Sampling is Nifty!
author: |
  Heather Gaya
date: "March 18, 2022"
output:
  beamer_presentation:
    highlight: tango
    theme: default
    includes:
      in_header: preamble.tex 
fig.align: center
fontsize: 9pt
---

```{r, setup, include=FALSE}
library(formatR)
library(styler)
library(jagsUI)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy = 'styler')
#knitr::opts_chunk$set(fig.width = 4.5, collapse = TRUE)
#knitr::opts_chunk$set(fig.height = 2.5, collapse = TRUE)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=40))
library(knitr)
library(runjags)
library(nimble)
library(coda)
library(plotrix)
```

# What is Distance Sampling?

 - We can never count every animal at a site, unless they are very very rare 
 
 \vfill
 
 - We know objects that are far away are harder to see than objects that are close by
 
  
 \vfill
 
 
 - Distance sampling uses this idea to help us estimate abundance and density 
  
 \vfill


# Distance Sampling Assumptions

Distance sampling makes some key assumptions!

\vfill

- Animals are distributed independent of lines (or points)

- On the line, detection is certain (100% detection at distance 0)

- Distances are recorded correctly

- Animals don't move before detection (and no double counting of individuals)

 

# An Example

Let's say we want to know abundance on a plot. Each dot is an animal.

```{r, echo =F}
set.seed(10)
x <- runif(250, 0,1)
y <- runif(250, 0,1)
transects <- data.frame(xstart = seq(.1, .9, by = .2), ystart = 0, xend = seq(.1, .9, by = .2), yend = 1 )
plot(x,y, main = "", xlab = "", ylab = "", xlim= c(-.1,1.1), ylim = c(-.1,1.1), pch = 19, cex = 1.1, xaxt = "n", yaxt = "n")
```

# An Example

We want to have even coverage of the plot, so we put down some transects and have our observers walk (or fly, or boat, or drive, or whatever) on them

```{r, echo =F}
plot(x,y, main = "", xlab = "", ylab = "", xlim= c(-.1,1.1), ylim = c(-.1,1.1), pch = 19, cex = 1.1, xaxt = "n", yaxt = "n")
segments(transects$xstart, transects$ystart, transects$xend, transects$yend)
```

```{r, echo = F,results = 'asis', fig.width= 7, fig.height= 5}
#library(animation)
set.seed(1)
  # how many animals are there to detect?
  # generate animals unformly over (0,1)x(0,1)
  xy.dat <- data.frame(x = x, y = y)
  # sort the points by y (bottom to top)
  # this is the direction in which we go along the transect
  xy.dat <- xy.dat[order(xy.dat$y),]
  # give each point an id
  xy.dat$id <- 1:(nrow(xy.dat))

  # list of transects
  n.transects <- nrow(transects)
  # vector to hold indices of detected points
  detected <- rep(FALSE,nrow(xy.dat))

  # vector to hold histogram of distances
  histdat <- c(0)

  # loop over transects
  k <- 0
  for(tr.id in 1:n.transects){
    # pick out the current transect
    this.tr <- transects[tr.id,]
    # calculate all distances to this transect first
    ddist <- abs(this.tr$xstart[1]-xy.dat$x)
    # loop over the possibly observed animals
    for(j in seq_along(xy.dat[,1])){
      # take this observation
      obs <- xy.dat[j,]
      sigma <- 0.02
      this.detected <- (exp(-ddist[j]^2/(2*sigma^2)) > runif(1))
      # maybe it was already detected? don't un-detect it!
      detected[obs$id] <- detected[obs$id] | this.detected
      # only do this animation stuff when we detect something
      if(this.detected){
        cat(paste0("# An Example \n"))
        # plot whole study area and all "animals"
        plot(xy.dat$x, xy.dat$y, xaxt = "n", yaxt = "n",
             xlab="", ylab="",
             main="",
             xlim=c(-0.01,1.01), ylim=c(-0.01,1.01),
             pch=20,cex=0.7,col=grey(0.7))
        segments(transects$xstart, transects$ystart, transects$xend, transects$yend, col = "grey")
        if(tr.id > 1){segments(transects$xstart[1:(tr.id-1)], transects$ystart[1:(tr.id-1)], transects$xend[1:(tr.id-1)], transects$yend[1:(tr.id-1)], col = "red")}

        #lapply(transects[1:tr.id,],lines, lty=2, col="grey")
        if(!is.null(detected)){
          # plot the animals that have been detected so far in blue
          points(xy.dat[detected,c("x","y")], pch=20,cex=0.7,col="blue")
        }
        # draw perpendicular line -- from transect to animal
        lines(x = c(this.tr$xstart[1], obs[1]),
              y = c(obs[2], obs[2]),col="red")
        # add this observation to the histogram
        histdat <- c(histdat,ddist[j])

        # draw "along transect" line showing how far we come so far
        lines(x=c(this.tr$xstart,this.tr$xend),y=c(this.tr$ystart[1],obs[2]),col="red")

        ## update the histogram plot
        xmax <- 3.5 * sigma
        # hist(histdat,ylim=c(0,35),xlim=c(0,xmax),
        #      main="Histogram of observed distances",xlab="Distance",
        #      breaks=seq(0,xmax,len=10))
        cat("\n\n---------------------\n\n")
      } # end of detected stuff
    } # end of current transect loop
    cat(paste0("# An Example \n"))
        # plot whole study area and all "animals"
        plot(xy.dat$x, xy.dat$y, xaxt = "n", yaxt = "n",
             xlab="", ylab="",
             main="",
             xlim=c(-0.01,1.01), ylim=c(-0.01,1.01),
             pch=20,cex=0.7,col=grey(0.7))
        segments(transects$xstart, transects$ystart, transects$xend, transects$yend, col = "grey")
        segments(transects$xstart[1:tr.id], transects$ystart[1:tr.id], transects$xend[1:tr.id], transects$yend[1:tr.id], col = "red")
        points(xy.dat[detected,c("x","y")], pch=20,cex=0.7,col="blue")
        if(tr.id != n.transects){cat("\n\n---------------------\n\n")}
  } # end loop over all transects
```

# An Example

You'll notice this sample isn't random - we clearly found more animals near the transects than far away.

```{r, echo =F}
hist(histdat,ylim=c(0,25),xlim=c(0,xmax),
             main="Histogram of Observed Distances",xlab="Distance",
             breaks=seq(0,xmax,len=10))
```

# The Math

If we can quantify this bias, we can estimate our "effective" search area, which gives us density.

\begin{equation*}
\begin{split}
D & = \frac{n}{a} \\
\\
N & = DA
\end{split}
\end{equation*}

where $n$ = number of animals in your searched area,

$N$ = true abundance (derived),

$A$ = total area of interest,

and $a$ = effective area searched


# The Math (Cont')

How do we estimate this? Integrals!

```{r, echo = F}

#plot(c(0,1), col = "white", axes = F, xlab = "", ylab = "", ylim = c(0,1), xlim = c(0,1))
#lines(c(.5,.5), c(0,1))
xs <- seq(0,1, by = .01)
p <- exp(-xs^2/(2*(.2)^2))
plot(xs, p, type = "l", lwd = 2, ylab = "Probability of Detection", xlab = "Distance from Transect", ylim = c(0, 1.1))
polygon(c(xs, 0), c(p, 0), col=rgb(red = 1, green = 0, blue =0, alpha = .5))
arrows(.6,.6 , .3,.25, length = .1 )
text(.65,.7, "Effective strip width")
text(.65,.85, bquote(integral(e^(-x^2/(2*sigma^2)))), cex = 1.5)

```

# Alternative Math

Alternatively, we can reformat the problem so that WE define the area and ask the model to tell us how many animals are in the area we choose. Our equation is now:

\begin{equation*}
D = \frac{N}{L(2W)}
\end{equation*}

where $N$ is the true abundance ($n$ + some unknown quantity of undetected animals)

$L$ is the length of transects walked

and $W$ is the width of interest (generally, 1/2 spacing between transects)


# Distance Sampling - Gopher Tortoises!

For my MS, we wanted to improve distance sampling techniques for Gopher Tortoise surveys throughout the Southeast.

Tortoise burrows make great sedentary objects for distance sampling!

\vfill

\begin{figure}
   \includegraphics[width=0.475\textwidth]{Adult.jpeg}
   \hfill
   \includegraphics[width=0.475\textwidth]{AdultBurrow.jpeg}
\end{figure}

# Even Clint Went Outside

Burrows can range in size quite dramatically, so we also wanted to include the problem of size variation in our model.

\vfill

\begin{figure}
   \includegraphics[width=0.475\textwidth]{tiny.jpeg}
   \hfill
   \includegraphics[width=0.475\textwidth]{clint.jpeg}
\end{figure}

# Gopher Tortoise Model Goals

We needed our model to do (4) things:

- Produce an accurate estimate of gopher tortoise abundance

- Allow for burrow size to influence detection probability

- Account for imperfect detection on the transect line

- Incorporate size-dependent burrow occupancy


# Formulating the Model
\large
Assuming some $M$ >> $N$:

\begin{align*}
{\color{blue} x_i} & \sim Uniform(0, {\color{teal}B}) & & \text{Distance to transect } \\
{\color{blue} w_i} & \sim Bern({\color{red} \psi}) & & \text{Is this tortoise real?}\\
{\color{red} \psi} & \sim Uniform(0, 1) & &\text{Only } \frac{N}{M} \text{ are real}\\
{\color{red} N} & = \sum_{i = 1}^{{\color{teal}M}} {\color{blue} w_i} & &\text{Count up how many burrows are real} \\
{\color{red}p_i} & = {\color{red}\xi_i} e^{\frac{-({\color{blue}x_i})^2}{2 {\color{red}\sigma_i}^2}} & & \text{Detection probability}\\
ln({\color{red}\sigma_i}) & = {\color{red}\beta_0} + {\color{red}\beta_1}{\color{blue} z_i} & & \text{Detection depends on size of burrow} \\
{\color{teal}y_i} & = Bern( {\color{red}p_i} {\color{blue}w_i}) & & \text{Was this burrow detected?} \\
{\color{blue}o_i} & \sim Bern({\color{red}\theta_i}) & & \text{Was this burrow occupied?} \\
logit({\color{red}\theta_i}) & = {\color{red}\alpha_0} + {\color{red}\alpha_1}{\color{blue} w_i} & & \text{smaller burrows more likely to  be occupied} \\
{\color{red} N_{occ}} & = \sum_{i = 1}^{{\color{teal}M}} {\color{blue} w_io_i} & &\text{Count up tortoises} \\
\end{align*}

# Variation in Detection

```{r, echo = F}
xs <- seq(0,1, by = .01)
p_1 <- exp(-xs^2/(2*(.3)^2))
p_2 <- .8*exp(-xs^2/(2*(.2)^2))
p_3 <- .5* exp(-xs^2/(2*(.05)^2))
plot(xs, p_1, type = "l", lwd = 2, ylab = "Probability of Detection", xlab = "Distance from Transect", ylim = c(0, 1.1), cex.axis = 2, cex.lab = 1.65)
lines(xs, p_2, col = "red")
lines(xs, p_3, col = "blue")
legend("topright",
       legend = sapply(c(.3,.2,0.05), function(x) as.expression(substitute(sigma == B,list(B = as.name(x))))),
      col = c("black", "red", "blue"), lty = 1, cex = 2)
#x = .4
points(c(.4, .4, .4), c(p_1[41],p_2[41], p_3[41]), pch = 19)
arrows(.5,.7,.4, p_1[41], length = .1, lty = 2)
text(.5,.8, "p(.4) = .41", cex = 2)
```

# Variation in Detection

```{r, echo = F}
plot(xs, p_1, type = "l", lwd = 2, ylab = "Probability of Detection", xlab = "Distance from Transect", ylim = c(0, 1.1), cex.axis = 2, cex.lab = 1.65)
lines(xs, p_2, col = "red")
lines(xs, p_3, col = "blue")
legend("topright",
       legend = sapply(c(.3,.2,0.05), function(x) as.expression(substitute(sigma == B,list(B = as.name(x))))),
      col = c("black", "red", "blue"), lty = 1, cex = 2)
#x = .4
points(c(.4, .4, .4), c(p_1[41],p_2[41], p_3[41]), pch = 19)
arrows(.5,.7,.4, p_1[41], length = .1, lty = 2)
text(.5,.8, "p(.4) = .41", cex = 2)
arrows(.5,.4,.4, p_2[41], length = .1, col = "red", lty = 2)
text(.55,.45, "p(.4) = .11", col = "red", cex = 2)
```

# Variation in Detection

```{r, echo = F}
plot(xs, p_1, type = "l", lwd = 2, ylab = "Probability of Detection", xlab = "Distance from Transect", ylim = c(0, 1.1), cex.axis = 2, cex.lab = 1.65)
lines(xs, p_2, col = "red")
lines(xs, p_3, col = "blue")
legend("topright",
       legend = sapply(c(.3,.2,0.05), function(x) as.expression(substitute(sigma == B,list(B = as.name(x))))),
      col = c("black", "red", "blue"), lty = 1, cex = 2)
#x = .4
points(c(.4, .4, .4), c(p_1[41],p_2[41], p_3[41]), pch = 19)
arrows(.5,.7,.4, p_1[41], length = .1, lty = 2)
text(.5,.8, "p(.4) = .41", cex = 2)
arrows(.5,.4,.4, p_2[41], length = .1, col = "red", lty = 2)
text(.55,.45, "p(.4) = .11", col = "red", cex = 2)
arrows(.7,.2,.4, p_3[41], length = .1, col = "blue", lty = 2)
text(.7,.25, "p(.4) = 0", col = "blue", cex = 2)
```

# Formulating the Model

Two small problems:

- How do we calculate detection on the line ($\xi_i$)?

- What sizes ($z_i$) were the burrows we didn't see?


\begin{figure}
   \includegraphics{SizeDist.jpeg}
\end{figure}


# Imperfect Detection on the Line
```{r, include = F}
defOut <- knitr::knit_hooks$get("plot")  # save the default plot hook
knitr::knit_hooks$set(plot = function(x, options) {  # set new plot hook ...
  x <- defOut(x, options)  # first apply the default hook
  if(!is.null(options$wrapfigure)) {  # then, if option wrapfigure is given ...
    # create the new opening string for the wrapfigure environment ...
    wf <- sprintf("\\begin{wrapfigure}{%s}{%g\\textwidth}", options$wrapfigure[[1]], options$wrapfigure[[2]])
    x  <- gsub("\\begin{figure}", wf, x, fixed = T)  # and replace the default one with it.
    x  <- gsub("{figure}", "{wrapfigure}", x, fixed = T)  # also replace the environment ending
  }
  return(x)
})
```
\large
Assuming a minimum burrow size of 4 cm wide and imperfect detection for all burrows $< b_{pt}$:
\begin{align*}
\xi_i & =
\begin{cases}
  \frac{1-b}{b_{pt}-4}(z_i-4) + b & 4 \leq z_i < b_{pt} \\
  1 & z_i \geq b_{pt} \\
\end{cases}
\\
b & \sim Uniform(0.2, 1) \\
b_{pt} & \sim Uniform(15, 20)
\end{align*}

```{r, echo = F,out.width = ".75\\textwidth", fig.align="center", wrapfigure = list("R", .9)}
plot(c(4, 20, 50), c(.2, 1, 1), lwd = 2, ylab = "Detection on the Line", xlab = "Size of Burrow", ylim = c(0, 1.1), cex.axis = 2, cex.lab = 1.65, type = "l", xlim = c(0, 50))
abline(v = 4, lty = 2)
text(2, .2, "b", cex = 2)
text(20, 1.1, "bpt", cex = 2)
points(c(4, 20),c(.2, 1), pch = 19)
```

# Dealing with Size

This part is extra tricky. Here's what a tortoise population size distribution might look like:

```{r, echo = F}
xx <- seq(4,65, by = .1)
a <- .8*dgamma(xx, shape = 70, rate = 1.6)
b <- .45*dgamma(xx, shape = 10, rate = .7)
c <- .3*dgamma(xx, shape = 27, rate = .9)
d <- .4*dgamma(xx, shape = 75, rate = 1.6)
plot(xx, a+b+c+d, type = "l", ylab = "", xlab = "Burrow Size", yaxt ="n", lwd =2)
```

# Dealing with Size

We chose to model this as a mixture of an unknown number of gamma distributions.

```{r, echo = F}
a <- .8*dgamma(xx, shape = 70, rate = 1.6)
b <- .45*dgamma(xx, shape = 10, rate = .7)
c <- .3*dgamma(xx, shape = 27, rate = .9)
d <- .4*dgamma(xx, shape = 75, rate = 1.6)
plot(xx, a+b+c+d, type = "l", ylab = "", xlab = "Burrow Size", yaxt ="n", lwd = 2)
lines(xx, a, col = "orange", lwd = 2)
lines(xx, b, col = "blue", lwd = 2)
lines(xx, c, col = "green", lwd = 2)
lines(xx, d, col = "purple", lwd =2)
```

# Dealing with Size

Each tortoise's size is drawn from a given gamma ($k$) with probability $p_k$.

\begin{align*}
P(z_i \in k) & \sim Dirichlet(\omega_1, \ldots, \omega_K) && \text{Choose a gamma} \\
\omega_k & \sim Gamma(.001, .001) && \text{Uniformative prior for $\omega$} \\
z_i & \sim Gamma(\alpha_k, \beta_k) && \text{Exact size depends on which gamma}
\end{align*}

# Model Summary

\begin{figure}
   \includegraphics{mod.jpeg}
\end{figure}

# Let's Put it in JAGS!
\tiny
*Note: full code w/ priors available on Github
```{r, eval = F, echo = T}
modelstring.Foo = "
model
{
	for (i in 1:(nind +nz)) {
		w[i] ~ dbern(psi)
		x[i] ~ dunif(0,Bx)
		z[i] ~ dgamma(a[i],c.beta[i])T(4,50)
		a[i] <- shape[clust[i]]
		c.beta[i] <- betaClust[clust[i]]
    clust[i] ~ dcat( pClust[1:Nclust] )
		sigma[i] <- exp(sigma.int+sigma.beta*z[i])
		logp[i] <- -((x[i]*x[i])/(2*sigma[i]*sigma[i]))
		p[i] <- exp(logp[i])*xi[i]
		xi[i] <- ifelse(z[i] < b.point, m*z[i]+intercept, 1)
		mu[i] <- w[i]*p[i]
		y[i] ~ dbern(mu[i])
    o[i]~ dbin(o2[i], 1)
    logit(o2[i]) <- o.int + z.beta*z[i]
    wo[i] <- o[i]*w[i]
	}
	intercept ~ dunif(.1,.8)
	b.point <- 20
	p.online <- m*4+intercept
	m <- (1-intercept)/b.point

	for (clustIdx in 1: Nclust) {
		shape[clustIdx] ~ dunif(1,100)
    betaClust[clustIdx] ~ dunif(.2,2)
		}
  pClust[1:Nclust] ~ ddirch(psizes)
	psi~ dunif(0,1)			#exists or not
	Nt <- sum(wo)
	N <- sum(w)
}
"
```

# Side Note

If we had ignored size variation and imperfect detection on the transect, our code would have been so much nicer...

```{r, eval = F, echo = T}
modelstring.Foo = "
model
{
	for (i in 1:(nind +nz)) {
		w[i] ~ dbern(psi)					# augmentation
		x[i] ~ dunif(0,Bx)				# distances
		logp[i] <- -((x[i]*x[i])/(2*sigma*sigma))
		p[i] <- exp(logp[i])
		mu[i] <- w[i]*p[i]
		y[i] ~ dbern(mu[i])
	}

	sigma ~ dunif(0, 20)
	psi~ dunif(0,1)			#exists or not

	N <- sum(w)
	D <- N/(2*L*Bx)			#burrow density
}
"
```

# Results

\begin{figure}
   \includegraphics[height=0.9\textheight]{Figure_Burrows.jpeg}
\end{figure}

# Results

\begin{figure}
   \includegraphics[height=0.9\textheight]{Figure_Burrows2.jpeg}
\end{figure}

# Making It More Complicated

This model works great for including individual covariates like size. But what about environmental covariates that affect detection?

\begin{figure}
   \includegraphics{fire.jpeg}
\end{figure}


# Making It More Complicated

Luckily, all we need to do is adjust our formula for $\sigma$ and choose a prior for vegetation ($v_i$).

\begin{align*}
{\color{red}p_i} & = {\color{red}\xi_i} e^{\frac{-({\color{blue}x_i})^2}{2 {\color{red}\sigma_i}^2}} & & \text{Detection probability}\\
ln({\color{red}\sigma_i}) & = {\color{red}\beta_0} + {\color{red}\beta_1}{\color{blue} z_i}+ {\color{red}\beta_2}{\color{blue} v_i} & & \text{Detection depends on size + veg} \\
{\color{teal}y_i} & = Bern( {\color{red}p_i} {\color{blue}w_i}) & & \text{Was this burrow detected?} \\
{\color{blue} v_i} & \sim Beta(d,e) & & \text{Prior for vegetation}
\end{align*}

# Testing it Out

In addition to simulation, we tested our new model on field data collected before and after prescribed burns.

\begin{figure}
   \includegraphics[height=0.8\textheight]{Veg3.jpeg}
\end{figure}

# Testing it Out

\begin{figure}
   \includegraphics[height=0.9\textheight]{Veg1.jpeg}
\end{figure}

# Testing it Out

\begin{figure}
   \includegraphics[height=0.9\textheight]{Veg2.jpeg}
\end{figure}

# Did we meet our goals?

We were able to produce a model that accounted for all (5) objectives:

- Produce an accurate estimate of gopher tortoise abundance

- Allow for burrow size to influence detection probability

- Account for imperfect detection on the transect line

- Incorporate size-dependent burrow occupancy

- Account for spatial heterogeneity in a detection covariate

# Pet Interlude

\begin{figure}
   \includegraphics[width=0.22\textwidth]{Opal.jpeg}
   \includegraphics[width=0.22\textwidth]{Cleo.jpeg}
   \includegraphics[width=0.22\textwidth]{Perrin.jpeg}
   \includegraphics[width=0.22\textwidth]{Arch.jpeg}
\end{figure}


# Current Bird Adventures

In my PhD, we also use distance sampling - but for birds!  Instead of transects we use point counts. 
\begin{figure}
\centering
\includegraphics[height=0.7\textheight]{cawa.jpeg}
\end{figure}

# Point Counts

Assuming that birds are uniformly placed, the distance from the observer might look something like this:

```{r, echo =F}
set.seed(2)
library(plotrix)
x <- runif(500, -2, 2)
y <- runif(500, -2, 2)
dist <- sqrt((x)^2 + (y)^2)
prob <- exp(-(dist^2)/(2*(.75^2)))
det <- rbinom(length(x), 1, prob)
seen.x <- x[det == 1]
seen.y <- y[det == 1]
seen.dist <- dist[det == 1]
```
```{r, echo =F, out.height= "75%"}
hist(dist, breaks = seq(0,3, by = .25), main = "", freq = F, yaxt = "n", xlab = "", ylab = "", col = "skyblue", cex.axis = 2)
```

# Point Counts

But the observed distance might look like this:
```{r, echo = F, out.height= "75%"}
hist(seen.dist, breaks = seq(0,3, by = .25), main = "", freq = F, yaxt = "n", xlab = "", ylab = "", col = "darkgreen", cex.axis = 2)
```

# Point Count Math

For point count data, we need to account for the increasing probability of being at a given distance as you move away from the observer. For a half-normal detection curve:

\begin{equation*}
p = \int 2\pi e^{\frac{-x^2}{2\sigma^2}} dx
\end{equation*}

When you convert this into JAGS code, you split the integral into distance bins of equal width (10m, 20m, etc). Then you can use this fancy integration-avoidance trick:

```{r, echo = T, eval = F}
pbar <- ((pnorm(b[j+1], 0, tau) - pnorm(b[j], 0, tau)) /
                  dnorm(0, 0, tau) / (b[j+1]-b[j]))*psi[j]
```

# Setup

In the field, we perform point counts across our study site in North Carolina. The goal is to estimate the yearly abundance of all bird species present in the basin. 

\includegraphics[height=0.75\textheight]{Site_Map.jpeg}

# Protocol

Instead of returning to these point counts multiple times, we combine distance sampling with time-to-first-detection methods to help us estimate detection. 
\vfill

Observers listen/look for birds for 10 minutes in 4 2.5-minute increments and record the estimated distance of each bird in each time period.

\vfill

This gives us 4 "visits" to work with - repeatability without all the effort! 

\vfill

During analysis, we run a dynamic N-mixture model (think Dail and Madsen 2011) to quantify the apparent survival, apparent emigration, and abundance of species at each site. 

# Fun Math

We group all detections into 10-m distance bins, $b$ and calculate all bin midpoints, $r_b$.The probability of a detection in distance bin $b$ with area $A_b$ for species $k$ at site $i$ in year $t$ is
\begin{equation*}
p_{iktb}^d = \frac{2\pi}{A_b} \int_{r_b}^{r_{b+1}} exp(\frac{-x^2_b}{2\sigma^2_{it}})dx
\end{equation*}

Yearly counts are then drawn from a binomial distribution given the latent site-abundance $N_{ikt}$, the availability probability (not described here), and the detection probability (distance + time-removal sampling). 

\begin{align*}
Dist_{ikt} & \sim Multi(\pi^d_{it1:10}, n_{ikt}) \\
n_{ikt} & \sim Binomial(N_{ikt}, \sum p^a_{it1:4} \sum p_{ikt1:10}^d)
\end{align*}


# Results so far

\includegraphics[height=0.75\textheight]{Abundance_All.jpeg}

# Results so far

\includegraphics[height=0.75\textheight]{Trailing_Predicted.jpeg}

# Results so far

\includegraphics[height=0.75\textheight]{Core_Predicted.jpeg}

# Results so far

\includegraphics[height=0.75\textheight]{SpeciesRichness.jpeg}

# Questions?


I probably ran out of time, but if not... questions?

\vfill

Feel free to email me at heather.gaya(@)uga.edu 

\vfill
Most of my code can be found at https://github.com/heathergaya or https://sites.google.com/view/heather-gaya/ 

\vfill

Also, while on the topic of self promotion, check out my LLC: https://www.hhecoanalytics.com/




